---
title: "Plotting and Interactive map"
author: "Rahima Baboo"
format: html
editor: source
output: rmarkdown::github_document
---

Clear my workspace

rm(list=ls())

Get working directory

getwd()

---
#Installing packages for data visualisation

```{r}
install.packages(c("tidyverse", "sf", "lwgeom", "terra", "stars", "exactextractr"))

install.packages(c("cowplot", "hrbrthemes", "knitr", "leaflet", "htmltools", "rosm", "ggspatial", "rnaturalearth", "mapview", "tmap", "leafpop"))

install.packages("prettymapr")
install.packages("webshot")

webshot::install_phantomjs()

```

---

# Calling libraries

library(tidyverse)
library(sf)
library(leaflet)
library(dplyr)
library(htmltools)
library(prettymapr)
library(rosm)
library(ggspatial)
library(leafpop)
library(mapview)
library(webshot)

---

#2020

##Reading and tidying data

```{r}
humpback_data_2020 <- read_delim("C:/Users/Rahima/OneDrive - University of Cape Town/Documents/Rahima postgrad/GIT repos/GIS-Project/Humpback data/Humpback data.csv", delim = ";") 

#Create a data frame
humpback_data_2020 <- as.data.frame(humpback_data_2020)

# Remove columns 15 to 20

humpback_data_2020 <- dplyr::select(humpback_data_2020, -c(14:19))

#select only 2020 data from this code
humpback_data_2020 <- humpback_data_2020 %>%
  filter(year == 2020, month %in% c(1, 2, 3, 11, 12))

#remove duplicate rows based on all columns
humpback_data_2020 <- humpback_data_2020 %>%
  distinct()

```

##converting data frame to spatial object

```{r}

library(mapview)
library(leafpop)

humpback_data_2020 <- st_as_sf(humpback_data_2020, coords = c("Longitude", "Latitude"), crs = 4326)

class(humpback_data_2020)

names(humpback_data_2020)
```
##Checking crs

```{r}
st_crs(humpback_data_2020)
class(humpback_data_2020)
```

##plot

```{r}
ggplot() + geom_sf(data=humpback_data_2020)
```

```{r}

whale_2020 <- ggplot() + 
  annotation_map_tile(type = "hotstyle", progress = "none") + 
  geom_sf(data = humpback_data_2020, color = "mediumorchid4", size = 2, show.legend = FALSE) + 
  labs(title = "Humpback Whale Sightings 2020") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  coord_sf(xlim = c(-18237508, 18237508), ylim = c(-17037508, 94688))  # Set global coordinate limits


plot(whale_2020)
```

##save image

```{r}
ggsave("whale_2020_plot.png", plot = whale_2020, width = 8, height = 4, dpi = 300)
```


#2021

##Reading and tidying data

```{r}
humpback_data_2021 <- read_delim("C:/Users/Rahima/OneDrive - University of Cape Town/Documents/Rahima postgrad/GIT repos/GIS-Project/Humpback data/Humpback data.csv", delim = ";") 

#Create a data frame
humpback_data_2021 <- as.data.frame(humpback_data_2021)

library(dplyr)

# Remove columns 14 to 20

humpback_data_2021 <- dplyr::select(humpback_data_2021, -c(14:19))

#select only 2020 data from this code
humpback_data_2021 <- humpback_data_2021 %>%
  filter(year == 2021, month %in% c(1, 2, 3, 11, 12))

#remove duplicate rows based on all columns
humpback_data_2021 <- humpback_data_2021 %>%
  distinct()

```

##converting data frame to spatial object

```{r}

library(mapview)
library(leafpop)

humpback_data_2021 <- st_as_sf(humpback_data_2021, coords = c("Longitude", "Latitude"), crs = 4326)

class(humpback_data_2021)

names(humpback_data_2021)
```

##Checking crs

```{r}
st_crs(humpback_data_2021)
class(humpback_data_2021)
```

##plot

```{r}
ggplot() + geom_sf(data=humpback_data_2021)
```

```{r}

whale_2021 <- ggplot() + 
  annotation_map_tile(type = "hotstyle", progress = "none") + 
  geom_sf(data = humpback_data_2021, color = "royalblue4", size = 2, show.legend = FALSE) + 
  labs(title = "Humpback Whale Sightings 2021") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  coord_sf(xlim = c(-18237508, 18237508), ylim = c(-17037508, 94688)) +
  coord_sf(xlim = c(-18237508, 18237508), ylim = c(-17037508, 94688))  # Set global coordinate limits

plot(whale_2021)
```

##save image

```{r}
ggsave("whale_2021_plot.png", plot = whale_2021, width = 8, height = 4, dpi = 300)
```


#2022

##Reading and tidying the data

```{r}
humpback_data_2022 <- read_delim("C:/Users/Rahima/OneDrive - University of Cape Town/Documents/Rahima postgrad/GIT repos/GIS-Project/Humpback data/Humpback data.csv", delim = ";") 

#Create a data frame
humpback_data_2022 <- as.data.frame(humpback_data_2022)

library(dplyr)

# Remove columns 14 to 20

humpback_data_2022 <- dplyr::select(humpback_data_2022, -c(14:19))

#select only 2020 data from this code
humpback_data_2022 <- humpback_data_2022 %>%
  filter(year == 2022, month %in% c(1, 2, 3, 11, 12))

#remove duplicate rows based on all columns
humpback_data_2022 <- humpback_data_2022 %>%
  distinct()

```

##converting data frame to spatial object

```{r}

library(mapview)
library(leafpop)

humpback_data_2022 <- st_as_sf(humpback_data_2022, coords = c("Longitude", "Latitude"), crs = 4326)

class(humpback_data_2022)

names(humpback_data_2022)
```

##Checking crs

```{r}
st_crs(humpback_data_2022)
class(humpback_data_2022)
```

##plot

```{r}
ggplot() + geom_sf(data=humpback_data_2022)
```

```{r}

whale_2022 <- ggplot() + 
  annotation_map_tile(type = "hotstyle", progress = "none") + 
  geom_sf(data = humpback_data_2022, color = "springgreen4", size = 2, show.legend = FALSE) + 
  labs(title = "Humpback Whale Sightings 2022") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  coord_sf(xlim = c(-18237508, 18237508), ylim = c(-17037508, 94688)) +
  coord_sf(xlim = c(-18237508, 18237508), ylim = c(-17037508, 94688))  # Set global coordinate limits

plot(whale_2022)

```

##save image

```{r}
ggsave("whale_2022_plot.png", plot = whale_2022, width = 8, height = 4, dpi = 300)
```


#2023

##Reading and tidying the data

```{r}
humpback_data_2023 <- read_delim("C:/Users/Rahima/OneDrive - University of Cape Town/Documents/Rahima postgrad/GIT repos/GIS-Project/Humpback data/Humpback data.csv", delim = ";") 

#Create a data frame
humpback_data_2023 <- as.data.frame(humpback_data_2023)

library(dplyr)

# Remove columns 14 to 20

humpback_data_2023 <- dplyr::select(humpback_data_2023, -c(14:19))

#select only 2020 data from this code
humpback_data_2023 <- humpback_data_2023 %>%
  filter(year == 2023, month %in% c(1, 2, 3, 11, 12))

#remove duplicate rows based on all columns
humpback_data_2023 <- humpback_data_2023 %>%
  distinct()

```

##converting data frame to spatial object

```{r}

library(mapview)
library(leafpop)

humpback_data_2023 <- st_as_sf(humpback_data_2023, coords = c("Longitude", "Latitude"), crs = 4326)

class(humpback_data_2023)

names(humpback_data_2023)
```

##Checking crs

```{r}
st_crs(humpback_data_2023)
class(humpback_data_2023)
```

##plot

```{r}
ggplot() + geom_sf(data=humpback_data_2023)

```

```{r}

whale_2023 <- ggplot() + 
  annotation_map_tile(type = "hotstyle", progress = "none") + 
  geom_sf(data = humpback_data_2023, color = "yellow1", size = 2, show.legend = FALSE) + 
  labs(title = "Humpback Whale Sightings 2023") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  coord_sf(xlim = c(-18237508, 18237508), ylim = c(-17037508, 94688)) +
  coord_sf(xlim = c(-18237508, 18237508), ylim = c(-17037508, 94688))  # Set global coordinate limits

plot(whale_2023)

```

##save image

```{r}
ggsave("whale_2023_plot.png", plot = whale_2023, width = 8, height = 4, dpi = 300)
```

---
#Interactive map 

##Default openstreet map tiles

```{r}

leaflet() %>%
  # Add default OpenStreetMap map tiles
  addTiles(group = "Default") %>%  
  # Add our points
  addCircleMarkers(data = humpback_data_2020,
                   group = "Humpback whales 2020",
                   radius = 3, 
                   color = "darkred") 

```

##Common sense checks with the data

```{r}

mapview(humpback_data_2020, 
        popup =  popupTable(humpback_data_2020,
            zcol = c("geometry", "ID")))

```

##Adding all data points to the map

```{r}
library(dplyr)

humpback_sightings <- bind_rows(humpback_data_2020, humpback_data_2021, humpback_data_2022, humpback_data_2023)

```

class(humpback_data_all)

```{r}

humpback_sightings$year <- as.factor(humpback_sightings$year)

int.map <- mapview(humpback_sightings, 
        zcol = "year",
        popup =  popupTable(humpback_sightings,
            zcol = c("geometry", "year")))

int.map

```

##save interactive map
```{r}

mapshot(int.map, file = "Humpback_whale_int.map.jpeg")

```